Data Analysis Extensions

### 2016-02-08 RFCs/HDF5_Library/AnalysisExt/analysis_ext-body.tex
Update metadata indexing design doc
Former-commit-id: b578bfebadc799e9faf5e2b0b5429a7c93f079de

The interface for metadata indexes in HDF5 containers is modeled after the interface 
for data element indexes in HDF5, with a plugin architecture that allows new metadata 
index packages to be developed and added at application runtime, without modifications 
to the core HDF5 library.
The HDF5 library can then interact with the third-party indexes created in this manner 
through a set of callback interfaces that present query information to the
indexing package and retrieve results from it.
An HDF5 metadata index can be considered as a generic key-value store, in the
sense that for each \textit{key} that represents an HDF5 object ID, is associated
either a link/attribute name or an attribute \textit{value}.

@@

The \texttt{create}, \texttt{remove}, \texttt{open}, \texttt{close} callbacks
are similar to the data index plugins, albeit they operate on a file location and
not on a single object. The \texttt{insert\_entry}, \texttt{remove\_entry} callbacks
respectively insert/remove a referenced object from the index. Values can be extracted from
the object ID, depending on the metadata type that the plugin will index.
When the \texttt{query} callback is invoked, a list of object IDs that matches
the query is returned. The \texttt{get\_size} callback returns the size of the index
and additional info callbacks may also be added depending on the needs.

@@

Are considered both a MDHIM and a B-Tree plugin.

@@

\section{Usage Example}

In the following example, we show how one can make use of the query and
indexing interface to retrieve a dataspace selection within a dataset.
For simplicity's sake, we first create a dataset within the file, then open it to
create and attach a new index, in order to finally query data from it.
Note that for convenience, calls to directly read data that corresponds to the
result of the index query may be moved to the high-level API in the future.

{\lstsetc
\begin{lstlisting}
#define NTUPLES 256

int
main(int argc, char *argv[])
{
    float data[NTUPLES];
    hsize_t dims[1] = {NTUPLES};
    hid_t t file_id, dataspace_id, dataset_id;
    hid_t query_id, result_space_id;
    size_t result_npoints;
    float *result;
    int i;

    /* Initialize data. */
    for(i = 0; i < NTUPLES; i++) data[i] = (float) i;

    /* Create file. */
    file_id = H5Fcreate(file_name, H5F_ACC_TRUNC, H5P_DEFAULT, H5P_DEFAULT);

    /* Create the data space for the dataset. */
    dataspace_id = H5Screate_simple(rank, dims, NULL);

    /* Create dataset. */
    dataset_id = H5Dcreate(file_id, "Pressure", H5T_NATIVE_FLOAT, dataspace_id,
        H5P_DEFAULT, H5P_DEFAULT, H5P_DEFAULT);

    /* Write dataset. */
    H5Dwrite(dataset_id, H5T_NATIVE_FLOAT, H5S_ALL, H5S_ALL, H5P_DEFAULT, data);

    /* Close the dataset. */
    H5Dclose(dataset_id);

    /* Close dataspace. */
    H5Sclose(dataspace_id);

    /* Open dataset. */
    dataset_id = H5Dopen(file_id, "Pressure", H5P_DEFAULT);

    /* Create index using FastBit. */
    H5Xcreate(file_id, H5X_PLUGIN_FASTBIT, dataset_id, H5P_DEFAULT);

    /* Close the dataset. */
    H5Dclose(dataset_id);

    /* Create a simple query */
    query_id = H5Qcreate(H5Q_TYPE_DATA_ELEM, H5Q_MATCH_EQUAL, H5T_NATIVE_FLOAT,
        &query_value);

    /* Open dataset. */
    dataset_id = H5Dopen(file_id, "Pressure", H5P_DEFAULT);

    /* Use query to select elements in the dataset. */
    result_space_id = H5Dquery(dataset_id, H5S_ALL, query_id, H5P_DEFAULT);

    /* Allocate space to read data. */ 
    result_npoints = (size_t) H5Sget_select_npoints(result_space_id);
    result = malloc(result_npoints * sizeof(float));

    /* Read data using result_space_id. */
    H5Dread(dataset_id, H5T_NATIVE_FLOAT, H5S_ALL, result_space_id,
        H5P_DEFAULT, result);

    /* Use result. */
    [...]

    /* Free result. */
    free(result);

    /* Close the dataset. */
    H5Dclose(dataset_id);

    /* Close dataspace. */
    H5Sclose(result_space_id);

    /* Close query. */
    H5Qclose(query_id);

    /* Close the file. */
    H5Fclose(file_id);
}
\end{lstlisting}
}

>>>>

The interface for metadata indexes in HDF5 containers is modeled after the interface 
for data element indexes in HDF5, with a plugin architecture that allows new metadata 
index packages to be developed and added at application runtime, without modifications 
to the core HDF5 library.
The HDF5 library can then interact with the third-party indexes created in this manner 
through a set of callback interfaces that present query information to the
indexing package and retrieve results from it.
An HDF5 metadata index can be considered as a generic key-value store, in the
sense that for each \textit{key} that can for instance represent either a
link/attribute name and/or an attribute \textit{value}, is associated an HDF5
object ID. However, this is up to the plugin to define that mapping and its
storage organization.

@@

The \texttt{create}, \texttt{remove}, \texttt{open}, \texttt{close} callbacks
are similar to the data index plugins, albeit they operate on a file location and
not on a single object. The \texttt{insert\_entry}, \texttt{remove\_entry} callbacks
respectively insert/remove a referenced object from the index. Information can
be directly extracted from the object ID, depending on the metadata type that
the plugin will choose to index.
When the \texttt{query} callback is invoked, a list of object IDs that matches
the query is returned. The \texttt{get\_size} callback\footnote{Additional info
callbacks may also be added depending on the needs.} returns the size of the index.

@@

Are considered both a MDHIM~\cite{Greenberg15} and a B-Tree based plugin to
provide metadata indexing capabilities. \textit{More details will be added
once the design for these plugins has been defined.}

@@

\section{Usage Examples}

These basic examples aim at showing the analysis components presented so far.
We describe in these examples the query and view creation process and how to
efficiently retrieve data from a file by adding indexes.

\subsection{Query and View}
In this example, we show how one can create a query, apply that query to a file,
and retrieve data from the resulting view.

{\lstsetc
\begin{lstlisting}
#define FILENAME "test.h5"
#define DATASETNAME "Pressure"

int
main(int argc, char *argv[])
{
    hid_t t file, query, subquery1, subquery2, view;
    float subquery2_value = 3.14;
    unsigned apply_result = 0;

    /* Create file */
    file = H5Fcreate(FILENAME, H5F_ACC_TRUNC, H5P_DEFAULT, H5P_DEFAULT);

    /* Create groups / datasets / attributes / etc */

    /* Close the file */
    H5Fclose(file);

    /* Create a simple query */
    subquery1 = H5Qcreate(H5Q_TYPE_LINK_NAME, H5Q_MATCH_EQUAL, DATASETNAME);
    subquery2 = H5Qcreate(H5Q_TYPE_DATA_ELEM, H5Q_MATCH_LESS_THAN,
        H5T_NATIVE_FLOAT, &subquery2_value);
    query = H5Qcombine(subquery1, H5Q_COMBINE_AND, subquery2);

    /* Open file */
    file = H5Fopen(FILENAME, H5F_ACC_RDONLY, H5P_DEFAULT);

    /* Apply query to file */
    view = H5Qapply(file, query, &apply_result, H5P_DEFAULT)

    /* Result should contain region reference */
    if (result & H5Q_REF_REG) {
        hid_t refs, ref_type, ref_space, space;
        size_t n_refs, ref_size;
        void *ref_buf;

        /* Read region reference from view */
        refs = H5Dopen(view, H5Q_VIEW_REF_REG_NAME, H5P_DEFAULT);
        ref_type = H5Dget_type(refs);
        ref_space = H5Dget_space(refs);
        n_refs = (size_t) H5Sget_select_npoints(ref_space);
        ref_size = H5Tget_size(ref_type);
        ref_buf = malloc(n_refs * ref_size)
        H5Dread(refs, ref_type, H5S_ALL, ref_space, H5P_DEFAULT, ref_buf);
        H5Dclose(refs);
        H5Sclose(ref_space);

        /* Get selection from region reference */
        space = H5Rget_region(file, (href_t *) ref_buf);

        /* Use selection result [...] */

        /* Close selection */
        H5Sclose(space);

        /* Free reference buffer */
        free(ref_buf);
    }

    /* Close view */
    H5Gclose(view);

    /* Close queries */
    H5Qclose(query);
    H5Qclose(subquery1);
    H5Qclose(subquery2);

    /* Close the file */
    H5Fclose(file);
}
\end{lstlisting}
}

\subsection{Data Index and Data Query}
In this example, we show how one can make use of the query and
indexing APIs to efficiently retrieve a dataspace selection within a dataset.

{\lstsetc
\begin{lstlisting}
#define NTUPLES 256
#define FILENAME "test.h5"
#define DATASETNAME "Pressure"

int
main(int argc, char *argv[])
{
    float data[NTUPLES], *result;
    hsize_t dims[1] = {NTUPLES};
    hid_t t file, dataspace, dataset, dcpl, query, result_space;
    size_t result_npoints;
    int i;

    /* Initialize data */
    for(i = 0; i < NTUPLES; i++) data[i] = (float) i;

    /* Create file */
    file = H5Fcreate(FILENAME, H5F_ACC_TRUNC, H5P_DEFAULT, H5P_DEFAULT);

    /* Create the data space for the dataset */
    dataspace = H5Screate_simple(1, dims, NULL);

    /* Create and set property list to use FastBit index */
    dcpl = H5Pcreate(H5P_DATASET_CREATE);
    H5Pset_index_plugin(dcpl, H5X_PLUGIN_FASTBIT);

    /* Create dataset */
    dataset = H5Dcreate(file, DATASETNAME, H5T_NATIVE_FLOAT, dataspace,
        H5P_DEFAULT, dcpl, H5P_DEFAULT);

    /* Write dataset */
    H5Dwrite(dataset, H5T_NATIVE_FLOAT, H5S_ALL, H5S_ALL, H5P_DEFAULT, data);

    /* Close the dataset */
    H5Dclose(dataset);

    /* Close dataspace */
    H5Sclose(dataspace);

    /* Close the property */
    H5Pclose(dcpl);

    /* Close the file */
    H5Fclose(file);

    /* Create a simple query */
    query = H5Qcreate(H5Q_TYPE_DATA_ELEM, H5Q_MATCH_EQUAL, H5T_NATIVE_FLOAT,
        &query_value);

    /* Open file */
    file = H5Fopen(FILENAME, H5F_ACC_RDONLY, H5P_DEFAULT);

    /* Open dataset */
    dataset = H5Dopen(file, DATASETNAME, H5P_DEFAULT);

    /* NB. Alternatively the index can be generated on the existing dataset.
     * This, however, requires the file to be open in H5F_ACC_RDWR so that
     * the index can be stored within the file and attached to the dataset.
     * H5Xcreate(dataset, H5X_PLUGIN_FASTBIT, H5P_DEFAULT); */

    /* Use query to select elements in the dataset */
    result_space = H5Dquery(dataset, H5S_ALL, query, H5P_DEFAULT);

    /* Allocate space to read data */ 
    result_npoints = (size_t) H5Sget_select_npoints(result_space);
    result = malloc(result_npoints * sizeof(float));

    /* Read data using result_space_id */
    H5Dread(dataset, H5T_NATIVE_FLOAT, H5S_ALL, result_space,
        H5P_DEFAULT, result);

    /* Use result [...] */

    /* Free result */
    free(result);

    /* Close the dataset */
    H5Dclose(dataset);

    /* Close dataspace */
    H5Sclose(result_space);

    /* Close query */
    H5Qclose(query);

    /* Close the file */
    H5Fclose(file);
}
\end{lstlisting}
}

\subsection{Metadata Index}
In this example, we show how one can make use of the query and
indexing APIs to efficiently execute a metadata query on a file.

{\lstsetc
\begin{lstlisting}
#define FILENAME "test.h5"
#define DATASETNAME "Pressure"

int
main(int argc, char *argv[])
{
    hid_t t file, fcpl, query, view;
    unsigned apply_result = 0;

    /* Create and set property list to use B-Tree index */
    fcpl = H5Pcreate(H5P_FILE_CREATE);
    H5Pset_index_plugin(fcpl, H5X_PLUGIN_BTREE);

    /* Create file */
    file = H5Fcreate(FILENAME, H5F_ACC_TRUNC, fcpl, H5P_DEFAULT);

    /* Create groups / datasets / attributes / etc */

    /* Close the property */
    H5Pclose(fcpl);

    /* Close the file */
    H5Fclose(file);

    /* Create a simple query */
    query = H5Qcreate(H5Q_TYPE_LINK_NAME, H5Q_MATCH_EQUAL, DATASETNAME);

    /* Open file */
    file = H5Fopen(FILENAME, H5F_ACC_RDONLY, H5P_DEFAULT);

    /* NB. Alternatively the index can be generated on the existing file.
     * This, however, requires the file to be open in H5F_ACC_RDWR so that
     * the index can be stored within the file and attached to it.
     * H5Xcreate(file, H5X_PLUGIN_BTREE, H5P_DEFAULT); */

    /* Apply query to file */
    view = H5Qapply(file, query, &apply_result, H5P_DEFAULT)

    /* Use view result [...] */

    /* Close view */
    H5Gclose(view);

    /* Close query */
    H5Qclose(query);

    /* Close the file */
    H5Fclose(file);
}
\end{lstlisting}
}

### 2016-02-08 RFCs/HDF5_Library/AnalysisExt/analysis_ext.tex
More update to analysis document
Former-commit-id: bc183e6cb18f23e94a8800a6dbc554da7d4c95b2
Update metadata indexing design doc
Former-commit-id: b578bfebadc799e9faf5e2b0b5429a7c93f079de

% Code snippets
\usepackage{listings}
\def\lstsetc{\lstset{language=C,
  %numbers=left,
  xleftmargin=1.8em,
  frame=L,
  framexleftmargin=0.2em,
  rulecolor=\color{rulecolor},
  numberstyle=\tiny\color{gray},
  stepnumber=1,
  showspaces=false, 
  showstringspaces=false,
  breaklines=true,
  basicstyle=\footnotesize\ttfamily,
  stringstyle=\itshape,
  commentstyle=\itshape\bfseries,
  morekeywords={main, size_t, malloc, free,
    hsize_t, hid_t, herr_t,
    H5X_class_t, H5X_type_t, H5X_info_t,
    H5Q_combine_op_t, H5Q_type_t, H5Q_match_op_t,
    H5Fcreate, H5Fclose,
    H5Dcreate, H5Dopen, H5Dclose, H5Dwrite, H5Dread, H5Dquery,
    H5Qcreate, H5Qclose, H5Qcombine, H5Qget_type, H5Qget_match_op,
    H5Qget_components, H5Qget_combine_op, H5Qencode, H5Qdecode,
    H5Qapply, H5Qapply_multi, H5Qapply_str, H5Qapply_dir,
    H5Screate_simple, H5Sclose, H5Sget_select_npoints,
    H5Xregister, H5Xunregister, H5Xcreate, H5Xremove, H5Xget_count, H5Xget_size,
    H5Xget_info
    }
  }
}

% Title, author, etc
\title{Data Analysis Extensions}
\author{Jerome Soumagne}
\author{Quincey Koziol}
\date{Feb 3, 2016}
\rfcversion{2014-07-17.v9}
\revision{Jul. 17, 2014}{Version 1 circulated for comment within The HDF Group.}
\revision{Aug. 20, 2014}{Version 2 circulated for comment within The HDF Group.}
\revision{Nov. 14, 2014}{Version 3 circulated for comment within The HDF Group.}
\revision{Feb. 25, 2015}{Version 4 edits and typo fixing.}
\revision{Jul. 24, 2015}{Version 5 edits to view object.}
\revision{Oct. 14, 2015}{Version 6 rework of query/view sections.}
\revision{Oct. 15, 2015}{Version 7 first draft of cross-container queries.}
\revision{Oct. 31, 2015}{Version 8 some corrections.}
\revision{Feb. 3, 2016}{Version 9 more text corrections.}

>>>>

% Code snippets
\usepackage{listings}
\def\lstsetc{\lstset{language=C,
  %numbers=left,
  xleftmargin=1.8em,
  frame=L,
  framexleftmargin=0.2em,
  rulecolor=\color{rulecolor},
  numberstyle=\tiny\color{gray},
  stepnumber=1,
  showspaces=false, 
  showstringspaces=false,
  breaklines=true,
  basicstyle=\footnotesize\ttfamily,
  stringstyle=\itshape,
  commentstyle=\itshape\bfseries,
  morekeywords={main, size_t, malloc, free,
    hsize_t, hid_t, herr_t,
    H5X_class_t, H5X_type_t, H5X_info_t,
    H5Q_combine_op_t, H5Q_type_t, H5Q_match_op_t,
    H5Fcreate, H5Fclose, H5Fopen,
    H5Gcreate, H5Gopen, H5Gclose,
    H5Dcreate, H5Dopen, H5Dclose, H5Dwrite, H5Dread, H5Dquery, H5Dget_type,
    H5Dget_space,
    H5Pcreate, H5Pclose, H5Pset_index_plugin,
    H5Qcreate, H5Qclose, H5Qcombine, H5Qget_type, H5Qget_match_op,
    H5Qget_components, H5Qget_combine_op, H5Qencode, H5Qdecode,
    H5Qapply, H5Qapply_multi, H5Qapply_str, H5Qapply_dir,
    H5Rget_region,
    H5Screate_simple, H5Sclose, H5Sget_select_npoints,
    H5Tget_size,
    H5Xregister, H5Xunregister, H5Xcreate, H5Xremove, H5Xget_count, H5Xget_size,
    H5Xget_info
    }
  }
}

% Title, author, etc
\title{Data Analysis Extensions}
\author{Jerome Soumagne}
\author{Quincey Koziol}
\date{Feb 8, 2016}
\rfcversion{2014-07-17.v10}
\revision{Jul. 17, 2014}{Version 1 circulated for comment within The HDF Group.}
\revision{Aug. 20, 2014}{Version 2 circulated for comment within The HDF Group.}
\revision{Nov. 14, 2014}{Version 3 circulated for comment within The HDF Group.}
\revision{Feb. 25, 2015}{Version 4 edits and typo fixing.}
\revision{Jul. 24, 2015}{Version 5 edits to view object.}
\revision{Oct. 14, 2015}{Version 6 rework of query/view sections.}
\revision{Oct. 15, 2015}{Version 7 first draft of cross-container queries.}
\revision{Oct. 31, 2015}{Version 8 some corrections.}
\revision{Feb. 3, 2016}{Version 9 more text corrections.}
\revision{Feb. 8, 2016}{Version 10 add metadata indexing and usage examples.}
